---
title: "Progress_Summarization"
output: pdf_document
date: "2025-02-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)

if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
library(dplyr)

if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}


if (!requireNamespace("tricycle", quietly = TRUE)) {
  BiocManager::install("tricycle")
}
library(tricycle)

if (!requireNamespace("org.Mm.eg.db", quietly = TRUE)) {
  BiocManager::install("org.Mm.eg.db")
}
library(org.Mm.eg.db)

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
library(ggplot2)

if (!requireNamespace("cowplot", quietly = TRUE)) {
  install.packages("cowplot")
}
library(cowplot)

if (!requireNamespace("foreach", quietly = TRUE)) {
  install.packages("foreach")
}
library(foreach)

if (!requireNamespace("scattermore", quietly = TRUE)) {
  install.packages("scattermore")
}
library(scattermore)

if (!requireNamespace("psych", quietly = TRUE)) {
  install.packages("psych")
}
library(psych)

library(parallel)

if (!requireNamespace("future.apply", quietly = TRUE)) {
  install.packages("future.apply")
}
library(future.apply)

if (!requireNamespace("reshape2", quietly = TRUE)) {
  install.packages("reshape2")
}
library(reshape2)
library(ggrepel)
library(tidyr)
library(shiny)
library(viridis)

# Set working directory
setwd("Your/Working/Directory/sc-DNA-Methylation-Cell-Cycle-Analysis")

# Read number of CpGs inside gene data
CpG_df <- read.csv("CpG_Num_Inside_Gene.csv", header = TRUE)
```

Function to expand a proportion to a binary vector of a given length
```{r, include = FALSE}
expand_to_cpg <- function(proportion, cpg_count) {
  # Calculate the number of 1s needed in the binary vector
  ones_count <- round(proportion * cpg_count)
  
  # Create a binary vector with the calculated number of 1s and 0s
  binary_vector <- rep(0, cpg_count)
  
  # Assign 1s to first coupse of positions of binary vector
  if (ones_count > 0) {
    binary_vector[1:ones_count] <- 1
  }
  
  return(binary_vector)
}
```

## Data Generation 
Generate data using cosine function and hyper parameters
```{r echo=FALSE, cars, include = FALSE}
###Simulation Setting cell-cycle
gene_num <- dim(CpG_df)[1]
set.seed(05092024)
x<-seq(0,2*pi,length.out=1000)
loc_pram<-runif(gene_num, min = 0, max = 2 * pi)
noise<-rnorm(n=1000,mean=0,sd=0.01)

####Generate two matrix
theta_matrix<-c()
for(i in 1:length(loc_pram)){
  gene_theta<-x-loc_pram[i]
  theta_matrix<-rbind(theta_matrix,gene_theta)
}
theta_matrix<-as.data.frame(theta_matrix)
rownames(theta_matrix)<-paste0('gene_',1:gene_num)
colnames(theta_matrix)<-paste0('cell_',1:1000)

m_matrix<-c()
for(i in 1:length(loc_pram)){
  gene_theta<-0.52*cos(x-loc_pram[i])+noise+0.5
  m_matrix<-rbind(m_matrix,gene_theta)
}
m_matrix<-as.data.frame(m_matrix)
rownames(m_matrix)<-paste0('gene_',1:gene_num)
colnames(m_matrix)<-paste0('cell_',1:1000)

```

Plot of generated data
```{r}
library(ggplot2)
library(ggrepel)
library(tidyr)
library(viridis)

# Make sure methylation proportion fall insde [0,1]
m_matrix[m_matrix < 0] <- 0
m_matrix[m_matrix > 1] <- 1


m_matrix_long <- as.data.frame(t(m_matrix))
m_matrix_long$cell <- 1:nrow(m_matrix_long)  
m_matrix_long <- pivot_longer(m_matrix_long, cols = starts_with("gene_"), names_to = "gene", values_to = "methylation")

# Select genes for plotting
selected_genes <- c("gene_288", "gene_4", "gene_3")

# Filter data for selected genes
plot_data <- m_matrix_long[m_matrix_long$gene %in% selected_genes, ]
cols_to_mark <- c(1, 250, 500, 750, 1000)
highlight_data <- plot_data[plot_data$cell %in% cols_to_mark, ]

color_palette <- brewer.pal(3, "Set2")

ggplot(plot_data, aes(x = cell, y = methylation, color = gene)) +
  geom_line(size = 0.5) +  
  geom_point(data = highlight_data, aes(x = cell, y = methylation, color = gene), size = 3) +  
  geom_text_repel(data = highlight_data, aes(label = paste0(" Cell ", cell)), size = 4, box.padding = 0.3) +
  geom_line(aes(x = cell, y = 0.52*cos((cell / 1000)*2*pi)+0.5), color = "gray30", linetype = "dashed", size = 1.5) +
  scale_color_manual(values = color_palette) +  
  theme_minimal(base_size = 14) +  
  labs(
       x = "Cell Cycle Progression",
       y = "Methylation Value",
       color = "Genes") +
  theme(
    legend.position = "top", 
    legend.title = element_text(size = 14), 
    legend.text = element_text(size = 12),
    panel.grid.major = element_line(color = "gray80", linetype = "dotted"),  
    panel.grid.minor = element_blank()
  )
```
## Data Processing

Expand to GcP level methylation data from gene-level
```{r}
n_genes <- 500
n_cells <- 1000

total_rows <- sum(CpG_df$total_cpg_num[1:n_genes])

expanded_matrix <- matrix(0, nrow = total_rows, ncol = n_cells)
row_names <- character(total_rows)

### Expand the Matrix
row_index <- 1  

for (gene in 1:n_genes) {
  cpg_count <- CpG_df$total_cpg_num[gene] 
  for (cell in 1:n_cells) {
    # Get the methylation proportion for this gene and cell
    proportion <- m_matrix[gene, cell]
    # Expand this proportion to a binary vector of length `cpg_count`
    binary_vector <- expand_to_cpg(proportion, cpg_count)
    # Assign this binary vector to the appropriate rows in the expanded matrix
    expanded_matrix[row_index:(row_index + cpg_count - 1), cell] <- binary_vector
  }
  for (cpg in 1:cpg_count) {
    row_names[row_index] <- paste0("g", gene, "CpG", cpg)
    # Move to the next block of rows in the expanded matrix
    row_index <- row_index + 1
  }
}

rownames(expanded_matrix)<-row_names
colnames(expanded_matrix)<-paste0('cell_',1:n_cells)

# Check number of Cells with 0 variance
dim(expanded_matrix[, apply(expanded_matrix, 2, var) > 0, drop = FALSE])

# Save expanded_matrix as csv for calculating Tetrachoric correlation in other scripts
write.csv(expanded_matrix, "Tetra_Cal/expanded.csv")
```

## Analysis Results
Running PCA on CpG-level methylation data and getting embedding space
```{r}
###PCA Analysis
M_cell_matrix<-t(expanded_matrix)
pca_res <- prcomp(M_cell_matrix, scale = TRUE)
```

Running PCoA on dissimilarity matrix of CpG-level methylation data constructed by column pairwise Tetrachoric correlation, and getting embedding space
```{r}
# tet <- tetrachoric(expanded_matrix)$rho
# Warning: Matrix was not positive definite, smoothing was done
tet <- read.csv("tet.csv")[,-1]
rownames(tet)<-colnames(m_matrix)
tet <- 1-abs(tet)

pcoa_result_tet <- cmdscale(tet, k = 2, eig = TRUE)
```

## Plot of Embedding Space

Plot of PCA embedding space with color gradient representing gene-specific methylation proportion
```{r}
g <- 500
temp_df <- data.frame(
  x = pca_res$x[, 1],
  y = pca_res$x[, 2],
  z = t(m_matrix[cpg,])
)

colnames(temp_df) <- c("x", "y", "gene")
highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_1","cell_250", "cell_500", "cell_750", "cell_1000"), ]

ggplot(temp_df, aes(x = x, y = y, color = gene)) +
  geom_point(size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(
    title = paste("Expression Pattern Over All Cells of Gene ", g, " in Embedding Space of PCA"),
    x = "Dimension 1",
    y = "Dimension 2",
    color = "Methylation Proportion"
  ) +
  geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
```

Plot of PCA embedding space with color gradient representing cell index order
```{r}
temp_df <- data.frame(
  x = pca_res$x[, 1],
  y = pca_res$x[, 2],
  z = seq(1:length(pca_res$x[, 2]))
)

colnames(temp_df) <- c("x", "y", "gene")
highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_1","cell_250", "cell_500", "cell_750", "cell_1000"), ]

ggplot(temp_df, aes(x = x, y = y, color = gene)) +
  geom_point(size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(
    x = "Dimension 1",
    y = "Dimension 2",
    color = "Index Position"
  ) +
  geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
```

Dynamic Plot of PCA embedding space with color gradient representing gene-specific methylation proportion varying different genes
```{r}
ui <- fluidPage(
  titlePanel("Dynamic Expression Pattern in PCA Embedding Space"),
  
  sidebarLayout(
    sidebarPanel(
      sliderInput("g", "Select Gene Index (g):", min = 1, max = n_genes, value = 500, step = 1)
    ),
    
    mainPanel(
      plotOutput("dynamicPCAPlot")
    )
  )
)

server <- function(input, output) {
  output$dynamicPCAPlot <- renderPlot({
    g <- input$g  
    
    temp_df <- data.frame(
      x = pca_res$x[, 1],
      y = pca_res$x[, 2],
      gene = t(m_matrix[g, ])
    )
    
    colnames(temp_df) <- c("x", "y", "gene")
    rownames(temp_df) <- paste0("cell_", 1:1000)  

    highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_1", "cell_250", "cell_500", "cell_750", "cell_1000"), ]
    
    ggplot(temp_df, aes(x = x, y = y, color = gene)) +
      geom_point(size = 1) +
      scale_color_gradient(low = "blue", high = "red") +
      theme_minimal() +
      labs(
        title = paste("Expression Pattern Over All Cells of Gene ", g, " in PCA Space"),
        x = "Dimension 1",
        y = "Dimension 2",
        color = "Methylation Proportion"
      ) +
      geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
  })
}

shinyApp(ui = ui, server = server)
```

Note that for PCoA plots, cell 1 to cell 500 will form a perfect circle and cell 500 to cell 1000 form another perfect circle. Thus for each plot with color gradient on gene-specific methylation proportion and cell index, we will provide plot cor cell 1 to cell 500 and cell 500 to cell 1000 seperately. 

Plot of PCoA embedding space with color gradient representing gene-specific methylation proportion for cell 1 to cell 500
```{r}
g <- 500

temp_df <- data.frame(
  x = pcoa_result_tet[["points"]][, 1][1:500],
  y = pcoa_result_tet[["points"]][, 2][1:500],
  z = t(m_matrix[g,][1:500])
)

colnames(temp_df) <- c("x", "y", "gene")
highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_1","cell_250", "cell_500"), ]

ggplot(temp_df, aes(x = x, y = y, color = gene)) +
  geom_point(size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(
    x = "Dimension 1",
    y = "Dimension 2",
    color = "Methylation Proportion"
  ) +
  geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
```

Plot of PCoA(Multi-Dimensional Scaling) embedding space with color gradient representing gene-specific methylation proportion for cell 500 to cell 1000.
```{r}
temp_df <- data.frame(
  x = pcoa_result_tet[["points"]][, 1][500:1000],
  y = pcoa_result_tet[["points"]][, 2][500:1000],
  z = t(m_matrix[g,][500:1000])
)

colnames(temp_df) <- c("x", "y", "gene")
highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_500","cell_750","cell_1000"), ]

ggplot(temp_df, aes(x = x, y = y, color = gene)) +
  geom_point(size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(
    x = "Dimension 1",
    y = "Dimension 2",
    color = "Methylation Proportion"
  ) +
  geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
```

Plot of PCoA(Multi-Dimensional Scaling) embedding space with color gradient representing cell index order for cell 1 to cell 500.
```{r}
temp_df <- data.frame(
  x = pcoa_result_tet[["points"]][, 1][1:500],
  y = pcoa_result_tet[["points"]][, 2][1:500],
  z = seq(1, length(pcoa_result_tet[["points"]][, 1][1:500]))
)

colnames(temp_df) <- c("x", "y", "gene")
highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_1","cell_250", "cell_500"), ]

ggplot(temp_df, aes(x = x, y = y, color = gene)) +
  geom_point(size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(
    x = "Dimension 1",
    y = "Dimension 2",
    color = "Index Position"
  ) +
  geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
```

Plot of PCoA(Multi-Dimensional Scaling) embedding space with color gradient representing cell index order for cell 500 to cell 1000.
```{r}
temp_df <- data.frame(
  x = pcoa_result_tet[["points"]][, 1][500:1000],
  y = pcoa_result_tet[["points"]][, 2][500:1000],
  z = seq(500, length(pcoa_result_tet[["points"]][, 1]))
)

colnames(temp_df) <- c("x", "y", "gene")

highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_500","cell_750", "cell_1000"), ]

ggplot(temp_df, aes(x = x, y = y, color = gene)) +
  geom_point(size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(
    x = "Dimension 1",
    y = "Dimension 2",
    color = "Index Position"
  ) +
  geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
```

Dynamic Plot of PCoA(Multi-Dimensional Scaling) embedding space with color gradient representing gene-specific methylation proportion varying different genes
```{r}
ui <- fluidPage(
  titlePanel("Dynamic Expression Pattern in PCoA Embedding Space"),
  
  sidebarLayout(
    sidebarPanel(
      sliderInput("g", "Select Gene Index (g):", min = 1, max = n_genes, value = 500, step = 1)
    ),
    
    mainPanel(
      plotOutput("dynamicPCoAPlot")
    )
  )
)

server <- function(input, output) {
  output$dynamicPCoAPlot <- renderPlot({
    g <- input$g  
    
    temp_df <- data.frame(
      x = pcoa_result_tet[["points"]][, 1],
      y = pcoa_result_tet[["points"]][, 2],
      gene = t(m_matrix[g, ])
    )
    
    colnames(temp_df) <- c("x", "y", "gene")
    rownames(temp_df) <- paste0("cell_", 1:1000)  

    highlight_obs <- temp_df[rownames(temp_df) %in% c("cell_1", "cell_250", "cell_500", "cell_750", "cell_1000"), ]
    
    ggplot(temp_df, aes(x = x, y = y, color = gene)) +
      geom_point(size = 1) +
      scale_color_gradient(low = "blue", high = "red") +
      theme_minimal() +
      labs(
        title = paste("Expression Pattern Over All Cells of Gene ", g, " in PCoA Space"),
        x = "Dimension 1",
        y = "Dimension 2",
        color = "Methylation Proportion"
      ) +
      geom_text(data = highlight_obs, aes(label = rownames(highlight_obs)), vjust = -1, hjust = 1)
  })
}

shinyApp(ui = ui, server = server)
```






